---
title: "Epiverse-TRACE Ecosystem Contributions"
subtitle: "Analysis of contributions to the broader R ecosystem during the Wellcome Epiverse-TRACE Phase 1 grant"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    fig-width: 10
    fig-height: 6
    self-contained: true
    embed-resources: true
execute:
  warning: false
  message: false
  echo: false
  freeze: true
editor: visual
---

```{r setup}
#| include: false
library(tidyverse)
library(DT)
library(plotly)
library(scales)
library(knitr)
library(kableExtra)
library(viridis)

# Load data
source_summary <- read_csv("data/contributions/source_summary_20250805_130519.csv")
contributor_summary <- read_csv("data/contributions/contributor_summary_20250805_130519.csv")
contributions_detail <- read_csv("data/contributions/epiverse_contributions_20250805_130519.csv")

# Clean and prepare data
source_summary <- source_summary %>%
  mutate(
    ecosystem_type = case_when(
      str_detect(repository, "^r-lib/") ~ "Core R Libraries (r-lib)",
      str_detect(repository, "^ropensci") ~ "ROpenSci",
      str_detect(repository, "^carpentries/") ~ "The Carpentries",
      TRUE ~ "Other Repositories"
    )
  )

contributor_summary <- contributor_summary %>%
  mutate(
    org_clean = case_when(
      org == "LSHTM/MRCG" ~ "LSHTM/MRCG",
      org == "data.org" ~ "data.org",
      org == "LSHTM" ~ "LSHTM",
      TRUE ~ "External"
    )
  )

# Calculate key metrics
total_contributions <- sum(source_summary$contributions)
total_repositories <- nrow(source_summary)
total_contributors <- nrow(contributor_summary)
core_ecosystem_contributions <- source_summary %>%
  filter(ecosystem_type %in% c("Core R Libraries (r-lib)", "ROpenSci", "The Carpentries")) %>%
  summarise(total = sum(contributions)) %>%
  pull(total)
core_ecosystem_pct <- round(core_ecosystem_contributions / total_contributions * 100)

top_3_contributors <- contributor_summary %>%
  slice_max(total_contributions, n = 3) %>%
  summarise(total = sum(total_contributions)) %>%
  pull(total)
top_3_pct <- round(top_3_contributors / total_contributions * 100)
```

::: {.callout-note icon="false"}
## Key Findings

**`r total_contributions` contributions** across **`r total_repositories` repositories** in the broader R ecosystem

**`r core_ecosystem_pct`%** of contributions were to critical ecosystem infrastructure: Core R libraries (r-lib), ROpenSci, and the Carpentries

**`r total_contributors` team members** contributed to the wider ecosystem at least once

**Top 3 contributors** (from data.org, LSHTM, and MRCG) accounted for **`r top_3_pct`%** of all contributions
:::

## Contribution Overview

### Total Contributions by Ecosystem Component

```{r ecosystem-overview}
ecosystem_summary <- source_summary %>%
  group_by(ecosystem_type) %>%
  summarise(
    repositories = n(),
    contributions = sum(contributions),
    contributors = sum(contributors),
    avg_stars = round(mean(stars, na.rm = TRUE)),
    total_stars = sum(stars, na.rm = TRUE)
  ) %>%
  arrange(desc(contributions))

p1 <- ecosystem_summary %>%
  ggplot(aes(x = reorder(ecosystem_type, contributions), y = contributions, fill = ecosystem_type)) +
  geom_col(alpha = 0.8) +
  geom_text(aes(label = contributions), hjust = -0.1, size = 4, fontface = "bold") +
  scale_fill_viridis_d(name = "Ecosystem Component") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  coord_flip() +
  labs(
    title = "Contributions by Ecosystem Component",
    subtitle = "Total number of contributions (issues + pull requests)",
    x = "",
    y = "Number of Contributions"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        plot.title = element_text(size = 14, face = "bold"),
        panel.grid.major.y = element_blank())

ggplotly(p1, tooltip = c("x", "y")) %>%
  layout(margin = list(l = 200))
```

### Repository Impact: Stars vs Contributions

```{r stars-contributions}
p2 <- source_summary %>%
  ggplot(aes(x = contributions, y = stars, color = ecosystem_type, size = contributors)) +
  geom_point(alpha = 0.7) +
  scale_color_viridis_d(name = "Ecosystem") +
  scale_size_continuous(name = "Contributors", range = c(2, 8)) +
  scale_x_log10(labels = comma_format()) +
  scale_y_log10(labels = comma_format()) +
  labs(
    title = "Community Impact vs Our Contributions",
    subtitle = "Repository stars (community adoption) vs our contributions",
    x = "Number of Contributions (log scale)",
    y = "GitHub Stars (log scale)"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(size = 14, face = "bold"))

ggplotly(p2, tooltip = c("colour", "x", "y", "size"))
```

------------------------------------------------------------------------

## Institutional Analysis

### Contributions by Organization

```{r org-analysis}
org_summary <- contributor_summary %>%
  group_by(org_clean) %>%
  summarise(
    contributors = n(),
    total_contributions = sum(total_contributions),
    avg_contributions = round(mean(total_contributions), 1),
    repositories = sum(repositories)
  ) %>%
  arrange(desc(total_contributions))

p3 <- org_summary %>%
  ggplot(aes(x = reorder(org_clean, total_contributions), y = total_contributions, fill = org_clean)) +
  geom_col(alpha = 0.8) +
  geom_text(aes(label = paste0(total_contributions, "\n(", contributors, " people)")),
            hjust = -0.1, size = 3.5) +
  scale_fill_viridis_d(name = "Organization") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  coord_flip() +
  labs(
    title = "Contributions by Institution",
    subtitle = "Total contributions and number of contributors per organization",
    x = "",
    y = "Total Contributions"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        plot.title = element_text(size = 14, face = "bold"),
        panel.grid.major.y = element_blank())

ggplotly(p3, tooltip = c("x", "y"))
```

### Top Contributors

```{r top-contributors}
top_contributors <- contributor_summary %>%
  slice_max(total_contributions, n = 10) %>%
  select(Organization = org_clean, Username = username,
         `Total Contributions` = total_contributions,
         Issues = issues, `Pull Requests` = pull_requests,
         `Repositories Contributed To` = repositories)

kable(top_contributors,
      caption = "Top 10 Contributors to the Broader Ecosystem") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(1:3, bold = TRUE, background = "#f0f8ff")
```

------------------------------------------------------------------------

## Contribution Timeline

```{r timeline}
# Process contribution timeline
timeline_data <- contributions_detail %>%
  mutate(
    created_date = as.Date(created_at),
    year_month = floor_date(created_date, "month")
  ) %>%
  filter(!is.na(created_date)) %>%
  count(year_month, type, name = "contributions") %>%
  complete(year_month, type, fill = list(contributions = 0))

p4 <- timeline_data %>%
  ggplot(aes(x = year_month, y = contributions, fill = type)) +
  geom_col(position = "stack", alpha = 0.8) +
  scale_fill_viridis_d(name = "Contribution Type",
                      labels = c("Issues", "Pull Requests")) +
  scale_x_date(date_labels = "%Y-%m", date_breaks = "3 months") +
  labs(
    title = "Contribution Timeline",
    subtitle = "Monthly contributions to the broader ecosystem",
    x = "Date",
    y = "Number of Contributions"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

ggplotly(p4, tooltip = c("x", "y", "fill"))
```

------------------------------------------------------------------------

## Ecosystem Contributions Summary

```{r summary-table}
kable(ecosystem_summary,
      caption = "Summary Statistics by Ecosystem Component",
      col.names = c("Ecosystem Component", "Repositories", "Contributions",
                   "Our Contributors", "Avg Stars", "Total Stars")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE) %>%
  column_spec(1, width = "30%") %>%
  row_spec(which(ecosystem_summary$ecosystem_type %in%
                c("Core R Libraries (r-lib)", "ROpenSci", "The Carpentries")),
           bold = TRUE, background = "#f0f8ff")
```

------------------------------------------------------------------------

<footer>*Dashboard generated on `r Sys.Date()` • Data collected on 2025-08-05*</footer>
